### 2010-03-12 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: ispell. Gosh.

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developped this technology in our
on-going work on Epigram~\cite{pigs:epigram}. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground-up.

To the reader new to type theory, this first section will provide a
gentle introduction to dependently-typed systems. To the reader
familiar with type theory, she will recognize the basis of her
favorite type theory, be it Agda or Coq for instance. This scene sets,
we have a common ground for discussion.

@@

In the following, we adopt a standard presentation for our type
theory. It is defined by three forms of judgements: context validity
judgements, typing judgements, and equality judgements. Formally,
these judgments are presented as follow:

@@

\begin{wstructure}
    <- Judgemental equality
        <- Presentation independant of particular implementation choice
        -> Model in Agda, intensional
        -> Used in Epigram, OTT
\end{wstructure}

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modeled in Agda, which
features an intensional equality. On the other hand, it is used in
Epigram, which equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of propositional or even
fully extensional systems to also find their way through this
presentation.

@@

The rules ensuring Context Validity (Fig.\ref{fig:context-validity})
are unsurprisingly common. They simply garantee the validity of the
empty context, as well as the linear extension of a context by a
well-typed variable.

@@

The typing judgements (Fig.\ref{fig:typing-judgements}) are just as
familiar. We recognize the standard presentation of $\Pi$ and $\Sigma$
types, respectively inhabited by lambda terms and dependent
pairs. Naturally, there are rules for function application and
projections of $\Sigma$-types. Equal types can be substituted, thanks
to the conversion rule.

@@

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. As one would expect,
judgemental equality is respects symmety, reflexivity, and
transitivity. Further, it also encompasses the $β$-rules for function
application and pairs. Finally, equality under lambda is dealt by the
$ξ$-rule.

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper, clearly specifying which extensions we
require.

@@

\begin{wstructure}
!!! Need Help !!!
<- Meta-theoretical properties
    <- Assuming a stratified discipline
    <> The point here is to reassert that dependent types are not evil, 
       there is no non-terminating type-checker, or such horrible lies <>
    -> Strongly normalizing
        -> Every program terminates
    -> Type-checking terminates
    ???
\end{wstructure}

This completes our presentation of the type theory. Assuming a
stratified discpline of universe, the system we have described enjoy
some very strong meta-theoretical properties. Unlike simply typed
languages, such as Haskell, dependently-typed systems are
\emph{strongly normalizing}: every program that type-checks
terminates. Moreover, type-checking is decidable and can therefore be
implemented by a terminating algorithm.

@@

Again, there is a clear equivalent in the full-$\Set$ world: function
application. The operational behavior of $\F{switch}$ is
straightforward: $x$ is peeled off as we move deeper inside the nested
tuple $b$. When $x$ equals $\Ze$, we simply return the value we are
pointing to.

@@

As we have seen, finite sets are not just a way of making
data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
come with a notion of finite function space. However, we had to
extract that intuition from the type, by a careful reading. At a first
glance, this seems to contradict our argument in favor of types for
coding representation. Here, we are overflown by low-level, very
precise type information.

@@

\begin{wstructure}
<- Formalization: type propagation
    <- Motivation
        -> High-level syntax
            -> exprIn: types are pushed in
                <- Subject to type *checking*
            -> exprEx: types are pulled from
                <- Subject to type *synthesis*
        -> Translated into our low-level type theory
        -> Presented as judgements
    -> Presentation mirors typing rule of [figure] 
        -> Ignore identitical judgements
\end{wstructure}

We adapt and formalize this \emph{type propagation} mechanism to our
richer type theory. First of all, we define a high-level syntax,
composed by \emph{expressions}. Expressions divided into two syntactic
categories: $\exprIn$ into which types are pushed in, and $\exprEx$
from which types are pulled out. Following the bidirectional
presentation, $\exprIn$ are subject to type \emph{checking}, while
$\exprEx$ are subject to type \emph{synthesis}.

@@

A second consequence concerns the definition of a finite set. We can
abstract away the details of code and use a familiar syntax
instead. Hence, we use the syntax $\Enum\: [t_1 \ldots t_n]$ to denote
the finite sets composed by the nammed elements $t_1$, \ldots,
$t_n$. The translation to elements of $\EnumU$ is straightforward.

@@

Finally, and more importantly, we are able to present our user an
abstraction of \emph{finite} function space. This work as follow. When
an expression $t$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realize our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

@@

\begin{wstructure}
<- Summary
    -> Not a novel technique [refs?]
        /> Used as a boilerplate scrapper
    -> Make dealing with codes *practical*
        <- Example: Finite sets/finite function space
        -> We should not restrain ourself in using codes
            <- We know how to present them to the user
-> Will extend this machinery in further sections
\end{wstructure}

In this section, we have developped a type propagation system, based
on bidirectional type-checking. Using bidirectional type-checking as a
boilerplate scrapper is a well-known technique~\cite{who?} \note{I
  would need some citation of people using bidirectional presentation
  of their type theory}. In our case, we have shown how to instrument
bidirectionality to rationalize the expressivity of dependent
types. We have illustrated our approach with finite sets. We have
successfully abstracted away the low-level presentation of finite
sets, offering a convenient syntax instead.

This example teaches us that we should not be afraid by codes, as far
as type information is available. We have shown how to rationalize
this information in a formal presentation. Hence, we have proved that
programming with such objects is practical. As we introduce more codes
in our theory, we will show how to extend the framework we have
developped so far.

>>>>

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developed this technology in our
on-going work on Epigram~\cite{pigs:epigram}. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground-up.

To the reader new to type theory, this first section will provide a
gentle introduction to dependently-typed systems. To the reader
familiar with type theory, she will recognise the basis of her
favourite type theory, be it Agda or Coq for instance. This scene sets,
we have a common ground for discussion.

@@

In the following, we adopt a standard presentation for our type
theory. It is defined by three forms of judgements: context validity
judgements, typing judgements, and equality judgements. Formally,
these judgements are presented as follow:

@@

\begin{wstructure}
    <- Judgemental equality
        <- Presentation independent of particular implementation choice
        -> Model in Agda, intentional
        -> Used in Epigram, OTT
\end{wstructure}

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modelled in Agda, which
features an in-tensional equality. On the other hand, it is used in
Epigram, which equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of fully extensional
systems to also find their way through this presentation.

@@

The rules ensuring Context Validity (Fig.\ref{fig:context-validity})
are unsurprisingly common. They simply guarantee the validity of the
empty context, as well as the linear extension of a context by a
well-typed variable.

@@

The typing judgements (Fig.\ref{fig:typing-judgements}) are just as
familiar. We recognise the standard presentation of $\Pi$ and $\Sigma$
types, respectively inhabited by lambda terms and dependent
pairs. Naturally, there are rules for function application and
projections of $\Sigma$-types. Equal types can be substituted, thanks
to the conversion rule.

@@

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. As one would expect,
judgemental equality respects symmetry, reflexivity, and
transitivity. Further, it also encompasses the $\beta$-rules for
function application and pairs. Finally, equality under lambda is
dealt with by the $\xi$-rule.

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper.

@@

\begin{wstructure}
!!! Need Help !!!
<- Meta-theoretical properties
    <- Assuming a stratified discipline
    <> The point here is to reassert that dependent types are not evil, 
       there is no non-terminating type-checker, or such horrible lies <>
    -> Strongly normalising
        -> Every program terminates
    -> Type-checking terminates
    ???
\end{wstructure}

This completes our presentation of the type theory. Assuming a
stratified discipline of universe, the system we have described enjoy
some very strong meta-theoretical properties. Unlike simply typed
languages, such as Haskell, dependently-typed systems are
\emph{strongly normalising}: every program that type-checks
terminates. Moreover, type-checking is decidable and can therefore be
implemented by a terminating algorithm.

@@

Again, there is a clear equivalent in the full-$\Set$ world: function
application. The operational behaviour of $\F{switch}$ is
straightforward: $x$ is peeled off as we move deeper inside the nested
tuple $b$. When $x$ equals $\Ze$, we simply return the value we are
pointing to.

@@

As we have seen, finite sets are not just a way of making
data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
come with a notion of finite function space. However, we had to
extract that intuition from the type, by a careful reading. At a first
glance, this seems to contradict our argument in favour of types for
coding representation. Here, we are overflown by low-level, very
precise type information.

@@

\begin{wstructure}
<- Formalisation: type propagation
    <- Motivation
        -> High-level syntax
            -> exprIn: types are pushed in
                <- Subject to type *checking*
            -> exprEx: types are pulled from
                <- Subject to type *synthesis*
        -> Translated into our low-level type theory
        -> Presented as judgements
    -> Presentation mirrors typing rule of [figure] 
        -> Ignore identical judgements
\end{wstructure}

We adapt and formalise this \emph{type propagation} mechanism to our
richer type theory. First of all, we define a high-level syntax,
composed by \emph{expressions}. Expressions divided into two syntactic
categories: $\exprIn$ into which types are pushed in, and $\exprEx$
from which types are pulled out. Following the bidirectional
presentation, $\exprIn$ are subject to type \emph{checking}, while
$\exprEx$ are subject to type \emph{synthesis}.

\note{Why don't we need a context validity judgement, again?}

@@

A second consequence concerns the definition of a finite set. We can
abstract away the details of code and use a familiar syntax
instead. Hence, we use the syntax $\Enum\: [t_1 \ldots t_n]$ to denote
the finite sets composed by the named elements $t_1$, \ldots,
$t_n$. The translation to elements of $\EnumU$ is straightforward.

@@

Finally, and more importantly, we are able to present our user an
abstraction of \emph{finite} function space. This work as follow. When
an expression $t$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realise our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

@@

\begin{wstructure}
<- Summary
    -> Not a novel technique [refs?]
        /> Used as a boilerplate scrapper
    -> Make dealing with codes *practical*
        <- Example: Finite sets/finite function space
        -> We should not restrain our self in using codes
            <- We know how to present them to the user
-> Will extend this machinery in further sections
\end{wstructure}

In this section, we have developed a type propagation system, based
on bidirectional type-checking. Using bidirectional type-checking as a
boilerplate scrapper is a well-known technique~\cite{who?} \note{I
  would need some citation of people using bidirectional presentation
  of their type theory}. In our case, we have shown how to instrument
bi-directionality to rationalise the expressivity of dependent
types. We have illustrated our approach with finite sets. We have
successfully abstracted away the low-level presentation of finite
sets, offering a convenient syntax instead.

This example teaches us that we should not be afraid by codes, as far
as type information is available. We have shown how to rationalise
this information in a formal presentation. Hence, we have proved that
programming with such objects is practical. As we introduce more codes
in our theory, we will show how to extend the framework we have
developed so far.
